A hazards approach to the biometric analysis of infant mortality

### 2023-03-20 paper2.Rmd
work on birth cohorts

```{r plotcohort, fig.cap = "By birth month, cohorts.", fig.height = 9, eval = FALSE}
oldpar <- par(mfrow = c(4, 3))
for (mon in levels(dat$bmonth)){
    drawBB2(dat[dat$bmonth == mon, ], main = mon, printLegend = FALSE)
}
par(oldpar)
```

```{r plotcohort2, fig.cap = "By birth season, cohorts.", fig.height = 8}
oldpar <- par(mfrow = c(2, 2))
for (seas in levels(dat$bseason)){
    drawBB2(dat[dat$bseason == seas, ], main = seas, printLegend = FALSE)
}
par(oldpar)
```

We see that the post-neonatal part of the total cumulative hazards are more or
less convex (bend upward at the end) for those born in spring or summer, while 
the curves are concave for births in the fall or winter. Bad fits to "constant hazard"
and "uniform distribution".

@@

```{r plotperiod, fig.cap = "By calendar month, periods.", fig.height = 9, eval = FALSE}
oldpar <- par(mfrow = c(4, 3))
for (mon in levels(dat$bmonth)){
    drawBB2(dat[dat$month == mon, ], main = mon, printLegend = FALSE)
}
par(oldpar)
```

```{r plotperiod2, fig.cap = "By calendar season, periods.", fig.height = 8}
oldpar <- par(mfrow = c(2, 2))
for (seas in levels(dat$season)){
    drawBB2(dat[dat$season == seas, ], main = seas, printLegend = FALSE)
}
par(oldpar)
```
The periods are of course more homogeneous regarding weather and external threats,
and it shows off in that in all four periods the suggested models fit well, here shown by the
"constant hazards model" for exogenous mortality.

>>>>

```{r plotcohort, fig.cap = "By birth month, cohorts, Västerbotten 1861-1950.", fig.height = 8, eval = TRUE}
oldpar <- par(mfrow = c(1, 1))
dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(bmonth), data = dato)
cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
haz <- hazards(fit)
pd <- numeric(12)
for(i in 1:12){
    haz[[i]] <- rbind(c(0, 0), haz[[i]])
    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
}
plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
     type = "l", xlab = "Days", ylab = "Cumulative hazards")
for (i in 2:12){
    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
}
legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
att <- c(0, 28, 183, 365)
axis(1, at = g(att), labels = att)
axis(2, las = 1)
box()
abline(v = g(28), lty = 3)
abline(h = 0)
fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
haza <- hazards(fita)
lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
par(oldpar)
```

```{r plotcohort2, fig.cap = "By birth season, cohorts.", fig.height = 8, eval = FALSE}
oldpar <- par(mfrow = c(2, 2))
for (seas in levels(dat$bseason)){
    drawBB2(dat[dat$bseason == seas, ], main = seas, printLegend = FALSE)
}
par(oldpar)
```

We see that the post-neonatal part of the total cumulative hazards are more or
less convex (bend upward at the end) for those born in spring (red) or summer (green), while 
the curves are concave for births in the winter (blue). Bad fits to "constant hazard"
and "uniform distribution". However, as the solid black curve shows the 12-month mixture 
has an excellent fit to a straight line.

Note the two outliers, February and October. Born in February means a tough first month of 
life, but surviving that, the future prospects are great. The opposite holds for those born in October.

@@

```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861-1950.", fig.height = 9, eval = TRUE}
oldpar <- par(mfrow = c(4, 3))
for (mon in levels(dat$bmonth)){
    drawBB2(dat[dat$month == mon, ], main = mon, printLegend = FALSE)
}
par(oldpar)
```

```{r plotperiod2, fig.cap = "By calendar season, periods.", fig.height = 8, eval = FALSE}
oldpar <- par(mfrow = c(2, 2))
for (seas in levels(dat$season)){
    drawBB2(dat[dat$season == seas, ], main = seas, printLegend = FALSE)
}
par(oldpar)
```
The periods are of course more homogeneous regarding weather and external threats,
and it shows off in that in all four periods the suggested models fit well, here shown by the
"constant hazards model" for exogenous mortality.
